Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
edits to abstract and intro

\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
% Anne: Added a lead-in motivator to address the submission as a Experience paper
% and tried to improve writing... ( I am assumeing no abstract word count limit...)
Challenges related to development, deployment, and maintenance of 
reusable software for science are becoming a growing concern. 
Research of many scientists now increasingly depends  on 
the quality and availability of software upon which their works are built. 
At the same time, software specialists have more than enough to do 
without having to ``reinvent the wheel.'' To highlight some of these 
issues and share experiences, the First Workshop on Sustainable 
Software for Science: Practice and Experiences (WSSSPE1) was held 
in November 2013 in conjunction with the Supercomputing Conference SC13.
The workshop featured keynote presentations and based its discussion 
on them, as well as on a large number (54) of solicited extended abstracts 
that were grouped into three themes presented via panels that encouraged 
audience participation. A set of collaborative notes was taken during the workshop.

This document summarizes 
the issues and shared experiences discussed at the workshop, including
cross-cutting issues and use cases. It joins a nascent
literature seeking to understand what drives software work in science,
and how it is impacted by the reward systems of science. 
These incentives can determine the extent to which
developers are motivated to build software for the long-term, for the
use of others, and whether to work collaboratively or separately. We 
also explore on community building, leadership, and dynamics in relation to
successful scientific software.

@@

@@

Because progress in scientific research is dependent on the quality of
and accessibility to software at all levels, it is now critical to
address many challenges related to the development, deployment, and
maintenance of reusable software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists, researchers, and
students are able to learn and adopt software-related skills and
methodologies. Established researchers are already acquiring some of
these skills, and in particular a specialized class of software
developers is emerging in academic environments as an integral part of
successful research teams. This first WSSSPE workshop provided a forum
for discussion of the challenges around sustaining scientific
software, including contributed short papers in the form of both
positions and experience reports. These short papers, as well as notes
from the debates around them, have been archived to provide a basis
for continued discussion, and have fed into the collaborative writing
of this report.  Some of the workshop submissions have been extended
to full papers, which form part of the same special journal edition in
which this report appears. The workshop generated a high level of
interest; an estimated~90 to~150 participants were in attendance at
different times of the day. The interest and discussions have already
led to follow-up activities: A smaller Python-specific workshop is
planned to be held at the~2014 SciPy conference, and a follow-on
WSSSPE2 workshop has been accepted for
the SC14 conference. In addition, funds to support the workshops have
been obtained from the US National Science Foundation (NSF) and the
Gordon and Betty Moore Foundation, and the original workshop website
has been turned into a community website to engender further
discussion and progress. Additionally, a minisymposium at the~2014
Society for Industrial and Applied Mathematics (SIAM) Annual Meeting
on ``Reliable Computational Science'' (SIAM AN14,
\url{http://meetings.siam.org}) is being co-organized by a WSSSPE1
participant to further explore some of the key issues raised at the
workshop.

This report attempts to summarize the various aspects of the workshop.
The remainder of this paper first gives an overview of the process
with which the workshop was put together (\S\ref{sec:process}), then
proceeds with summaries of the two keynote
presentations~(\S\ref{sec:keynotes}), followed by summaries of the
workshop papers grouped by the three thematic workshop panels to which
they were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Three
broader cross-cutting issues surfaced repeatedly, and are discussed
separately~(\S\ref{sec:cross-cutting}), as are use cases for
sustainable software~(\S\ref{sec:use-cases}). The summaries are based
not only on the papers and panel presentations, but also on the many
comments raised in both the onsite and online discussions that
accompanied the workshop elements, as documented by collaborative note
taking during the workshop~\cite{WSSSPE1-google-notes}. We conclude
with issues and lessons learned, as well as plans for future
activities~(\S\ref{sec:conclusions}).  The short papers accepted to
the workshop are listed in Appendix~\ref{sec:papers}, and a partial
list of workshop attendees can be found in
Appendix~\ref{sec:attendees}. The original call for papers is included in 
Appendix~\ref{sec:cfp}.

@@

In roughly following the aforementioned scientific software ecosystem
challenges, the accepted submissions were grouped into three main
categories, namely \emph{Developing and Supporting Software},
\emph{Policy}, and \emph{Communities}. Each category was assigned a
panel, with three to four panelists drawn from authors of the
associated submissions, who were each assigned to read and summarily
present a subset of the papers associated with the panel. The process
from organizing and advertising the workshop, to collecting and
reviewing the papers, and putting together the agenda was documented
by the organizers in a report~\cite{WSSSPE1-pre-report}, which they
self-archived in the same way as contributed papers.

For each panel, the paper summaries were followed by active discussion
involving panelists, onsite attendees, and often enough online
participants. The latter was facilitated by having a shared Google
Doc~\cite{WSSSPE1-google-notes} for collaborative note taking. Some of
the online discussion also took place on Twitter (hashtag
\texttt{\#wssspe}).

The workshop itself consisted of two keynote presentations followed by
the three panel sessions.  Remote participation was facilitated by a
live-cast of all keynotes and panels on Ustream.tv
(\url{http://ustream.tv}).

>>>>

\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
% Anne: Added a lead-in motivator to address the submission as a Experience paper
% and tried to improve writing... ( I am assumeing no abstract word count limit...)
Challenges related to development, deployment, and maintenance of 
reusable software for science are becoming a growing concern. 
Many scientists' research increasingly depends  on 
the quality and availability of software upon which their works are built. 
To highlight some of these 
issues and share experiences, the First Workshop on Sustainable 
Software for Science: Practice and Experiences (WSSSPE1) was held 
in November 2013 in conjunction with the SC13 Conference.
The workshop featured keynote presentations and a large number (54)
of solicited extended abstracts 
that were grouped into three themes and presented via panels.
A set of collaborative notes of the presentations and discussion
was taken during the workshop.

@@

This paper summarizes 
the issues and shared experiences that were discussed, including
cross-cutting issues and use cases. It joins a nascent
literature seeking to understand what drives software work in science,
and how it is impacted by the reward systems of science. 
These incentives can determine the extent to which
developers are motivated to build software for the long-term, for the
use of others, and whether to work collaboratively or separately. It 
also explores community building, leadership, and dynamics in relation to
successful scientific software.

@@

Because progress in scientific research is dependent on the quality of
and accessibility to software at all levels, it is now critical to
address many challenges related to the development, deployment, and
maintenance of reusable software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists, researchers, and
students are able to learn and adopt software-related skills and
methodologies. Established researchers are already acquiring some of
these skills, and in particular a specialized class of software
developers is emerging in academic environments as an integral part of
successful research teams. This first WSSSPE workshop provided a forum
for discussion of the challenges around sustaining scientific
software, including contributed short papers in the form of both
positions and experience reports. These short papers, as well as notes
from the debates around them, have been archived to provide a basis
for continued discussion, and have fed into the collaborative writing
of this report.  Some of the workshop submissions have been extended
to full papers, which form part of the same special journal edition in
which this report appears. The workshop generated a high level of
interest; an estimated~90 to~150 participants were in attendance at
different times of the day. The interest and discussions have already
led to follow-up activities: A smaller Python-specific workshop is
planned to be held at the~2014 SciPy conference, and a follow-on
WSSSPE2 workshop has been accepted for
the SC14 conference. In addition, funds to support the workshops have
been obtained from the US National Science Foundation (NSF) and the
Gordon and Betty Moore Foundation, and the original workshop website
has been turned into a community website to engender further
discussion and progress. Additionally, a minisymposium at the 2014
Society for Industrial and Applied Mathematics (SIAM) Annual Meeting
on ``Reliable Computational Science'' (SIAM AN14,
\url{http://meetings.siam.org}) is being co-organized by a WSSSPE1
participant to further explore some of the key issues raised at the
workshop.

This report attempts to summarize the various aspects of the workshop.
The remainder of this paper first gives an overview of the process
with which the workshop was put together (\S\ref{sec:process}), then
proceeds with summaries of the two keynote
presentations~(\S\ref{sec:keynotes}), followed by summaries of the
workshop papers grouped by the three thematic workshop panels to which
they were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Three
broader cross-cutting issues surfaced repeatedly, and are discussed
separately~(\S\ref{sec:cross-cutting}), as are use cases for
sustainable software~(\S\ref{sec:use-cases}). The summaries are based
not only on the papers and panel presentations, but also on the many
comments raised in both the onsite and online discussions that
accompanied the workshop elements, as documented by collaborative note
taking during the workshop~\cite{WSSSPE1-google-notes}. We conclude
with issues and lessons learned, as well as plans for future
activities~(\S\ref{sec:conclusions}).  The short papers accepted to
the workshop are listed in Appendix~\ref{sec:papers}, part of he original
call for papers is included in 
Appendix~\ref{sec:cfp}, and a partial
list of workshop attendees can be found in
Appendix~\ref{sec:attendees}.

@@

Roughly following the call for papers topics, the accepted submissions
were grouped into three main
categories, namely \emph{Developing and Supporting Software},
\emph{Policy}, and \emph{Communities}. Each category was assigned a
panel, with three to four panelists drawn from authors of the
associated submissions, who were each assigned to read and summarily
present a subset of the papers associated with the panel. The process
from organizing and advertising the workshop, to collecting and
reviewing the papers, and putting together the agenda was documented
by the organizers in a report~\cite{WSSSPE1-pre-report}, which they
self-archived in the same way as contributed papers.

The workshop also included two keynote presentations.
Remote participation was facilitated by a
live-cast of keynotes and panels via Ustream.tv
(\url{http://ustream.tv}).
In each panel, the paper summary presentations were followed by active discussion
involving panelists, onsite attendees, and often online
participants. The latter was facilitated by having a shared Google
Doc~\cite{WSSSPE1-google-notes} for collaborative note taking. Some of
the online discussion also took place on Twitter (hashtag
\texttt{\#wssspe}).
